Associations between Chlorophyll \emph{a} and various Microcystin-LR Health Advisory Concentrations

### 2015-09-25 manuscript/manuscript.Rmd
started confusion matrix

```{r setup, include=FALSE, echo=FALSE}
#Put whatever you normally put in a setup chunk.
#I usually at least include:
################################################################################
#Load packages
################################################################################
if(!require(microcystinchla)){
  devtools::install_github("jhollist/microcystinchla",
                           auth_token = getOption("Github_Access"))
}
if(!require(viridis)){
  install.pacakges("viridis") 
}
library("microcystinchla")
library("dplyr")
library("magrittr")
library("pander")
library("knitr")
opts_chunk$set(dev = 'pdf', fig.width=6, fig.height=5)

@@

#who_rec_low2_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
#                     xImpair = log10(mc_guides[5]), ProbComp = "gt", Exceed = "gte",
#                     ci = TRUE, R = 100, xW=wgt)

#mc_chla[5] <- get_chla(who_rec_low2_cp,0.5)

#who_rec_med1_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
#                     xImpair = log10(mc_guides[6]), ProbComp = "gt", Exceed = "gte",
#                     ci = TRUE, R = 100, xW=wgt)

#mc_chla[6] <- get_chla(who_rec_med1_cp,0.5)

#who_rec_med2_cp <- condprob(xX = nla_dat$log_chla,xY = nla_dat$log_micro, 
#                     xImpair = log10(mc_guides[7]), ProbComp = "gt", Exceed = "gte",
#                     ci = TRUE, R = 100, xW=wgt)

>>>>

```{r setup, include=FALSE, echo=FALSE}
#Put whatever you normally put in a setup chunk.
#I usually at least include:
################################################################################
#Load packages
################################################################################
if(!require(microcystinchla)){
  devtools::install_github("jhollist/microcystinchla",
                           auth_token = getOption("Github_Access"))
}
if(!require(viridis)){
  install.pacakges("viridis") 
}
library("microcystinchla")
library("dplyr")
library("magrittr")
library("pander")
library("knitr")
library("e1071")
opts_chunk$set(dev = 'pdf', fig.width=6, fig.height=5)

@@

#Agreement Tables
child_conmat<-table(chla=nla_dat$CHLA>=mc_chla[1],mcsyt=nla_dat$MCYST_TL_UGL>=0.3)
who_drink_conmat<-table(nla_dat$CHLA>=mc_chla[2],nla_dat$MCYST_TL_UGL>=1)
adult_conmat<-table(nla_dat$CHLA>=mc_chla[3],nla_dat$MCYST_TL_UGL>=1.6)
who_rec_conmat<-table(nla_dat$CHLA>=mc_chla[4],nla_dat$MCYST_TL_UGL>=2)
#Want to minimize false negatives.  Ratio or FN/FP+TP+TN

### 2015-09-25 manuscript/manuscript.tex
started confusion matrix

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use. However, compared to other water
quality measures, field measurements of microcystin are not commonly
available due to cost and require advanced understanding to interpret
results. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper we use this association to provide estimates
of chlorophyll \emph{a} that if exceeded would be indicative of a higher
probability of exceeding select health advisory concentrations for
microcystin-LR. Using the 2007 National Lakes Assessment and a
conditional probability approach that has been used in other water
quality settings, we identify chlorophyll \emph{a} concentrations that
are more likely than not to be associated with an exceedance of a
microcystin health advisory level. We look at the recent US EPA
standards for drinking water as well as the World Health Organization
levels for drinking water and recreational use. For microcystin
concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll \emph{a}
concentrations of 23.68, 63.94, 89.71, and 102.6, respectively. When
managing for these various microcystin levels exceeding these reported
chlorophyll \emph{a} concentrations should be a trigger for further
testing and possible management action.

@@

In the 2007 NLA, microcystin-LR concentrations ranged from 0.05 to 225.
Microcystin-LR concentrations of 0.05 \(\mu\)g/L represent the detection
limits. Any value greater than that indicates the presence of
microcystin-LR. Of those lakes with microcystin, the median
concentration was 0.51 and the mean was 0.51. Lastly, of all lakes
sampled, 21\% of lakes exceeded the U.S. EPA childrens drinking water
standard, 8.8\% of lakes exceeded the U.S. EPA adult drinking water
standard,11.7\% of lakes exceeded the WHO drinking water standard,and
7.3\% of lakes exceeded the WHO recreational standard. For chlorophyll
\emph{a}, the range was 0.07 to 936. All lakes had reported chlorophyll
\emph{a} concentrations that exceeded detection limits. The median
concentration was 7.79 and the mean was 29.6301946. The associations
between chlorophyll \emph{a} and the upper confidence interval with a
conditional probability of 50\% were 23.68, 63.94, 89.71, and 102.6 for
0.3, 1.0, 1.6 and 2.0 \(\mu\)g/L, respectively (Table
\ref{tab:mc_chla_table} \& Figure \ref{fig:multi_cp_plot}).

@@

This is the case as the probability of exceeding each of the four tested
health advisory levels increases as a function of chlorophyll \emph{a}
concentration (Figure \ref{fig:multi_cp_plot}). We use this association
to identify chlorophyll \emph{a} concentrations that are associated with
greater than even odds of exceeding a given health advisory level (Table
\ref{tab:mc_chla_table}). These represent 27.8\%, 11.9\%, 8.3\%, and
7.4\% of sample lakes for the U.S EPA childrens drinking water, the WHO
drinking water, the U.S. EPA Adult drinking water, and the WHO
recreational standards, respecitvely.

@@

\begin{longtable}[c]{@{}llrr@{}}
\caption{Chlorophyll \textit{a} concentrations that are associated with
a 50\% probability of exceeding a microcystin-LR health advisory
concentration. \label{tab:mc_chla_table}}\tabularnewline
\toprule
Source & Type & Microcystin & Chlorophyll\tabularnewline
\midrule
\endfirsthead
\toprule
Source & Type & Microcystin & Chlorophyll\tabularnewline
\midrule
\endhead
U.S. EPA & Drinking & 0.3 & 23.68\tabularnewline
WHO & Drinking & 1.0 & 63.94\tabularnewline
U.S. EPA & Drinking & 1.6 & 89.71\tabularnewline
WHO & Recreational & 2.0 & 102.60\tabularnewline
\bottomrule
\end{longtable}

>>>>

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use. However, compared to other water
quality measures, field measurements of microcystin are not commonly
available due to cost and require advanced understanding to interpret
results. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper we use this association to provide estimates
of chlorophyll \emph{a} that if exceeded would be indicative of a higher
probability of exceeding select health advisory concentrations for
microcystin-LR. Using the 2007 National Lakes Assessment and a
conditional probability approach that has been used in other water
quality settings, we identify chlorophyll \emph{a} concentrations that
are more likely than not to be associated with an exceedance of a
microcystin health advisory level. We look at the recent US EPA
standards for drinking water as well as the World Health Organization
levels for drinking water and recreational use. For microcystin
concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll \emph{a}
concentrations of 21.92, 65.2, 84.24, and 113.14, respectively. When
managing for these various microcystin levels exceeding these reported
chlorophyll \emph{a} concentrations should be a trigger for further
testing and possible management action.

@@

In the 2007 NLA, microcystin-LR concentrations ranged from 0.05 to 225.
Microcystin-LR concentrations of 0.05 \(\mu\)g/L represent the detection
limits. Any value greater than that indicates the presence of
microcystin-LR. Of those lakes with microcystin, the median
concentration was 0.51 and the mean was 0.51. Lastly, of all lakes
sampled, 21\% of lakes exceeded the U.S. EPA childrens drinking water
standard, 8.8\% of lakes exceeded the U.S. EPA adult drinking water
standard,11.7\% of lakes exceeded the WHO drinking water standard,and
7.3\% of lakes exceeded the WHO recreational standard. For chlorophyll
\emph{a}, the range was 0.07 to 936. All lakes had reported chlorophyll
\emph{a} concentrations that exceeded detection limits. The median
concentration was 7.79 and the mean was 29.6301946. The associations
between chlorophyll \emph{a} and the upper confidence interval with a
conditional probability of 50\% were 21.92, 65.2, 84.24, and 113.14 for
0.3, 1.0, 1.6 and 2.0 \(\mu\)g/L, respectively (Table
\ref{tab:mc_chla_table} \& Figure \ref{fig:multi_cp_plot}).

@@

This is the case as the probability of exceeding each of the four tested
health advisory levels increases as a function of chlorophyll \emph{a}
concentration (Figure \ref{fig:multi_cp_plot}). We use this association
to identify chlorophyll \emph{a} concentrations that are associated with
greater than even odds of exceeding a given health advisory level (Table
\ref{tab:mc_chla_table}). These represent 29.2\%, 11.7\%, 8.8\%, and
6.3\% of sample lakes for the U.S EPA childrens drinking water, the WHO
drinking water, the U.S. EPA Adult drinking water, and the WHO
recreational standards, respecitvely.

@@

\begin{longtable}[c]{@{}llrr@{}}
\caption{Chlorophyll \textit{a} concentrations that are associated with
a 50\% probability of exceeding a microcystin-LR health advisory
concentration. \label{tab:mc_chla_table}}\tabularnewline
\toprule
Source & Type & Microcystin & Chlorophyll\tabularnewline
\midrule
\endfirsthead
\toprule
Source & Type & Microcystin & Chlorophyll\tabularnewline
\midrule
\endhead
U.S. EPA & Drinking & 0.3 & 21.92\tabularnewline
WHO & Drinking & 1.0 & 65.20\tabularnewline
U.S. EPA & Drinking & 1.6 & 84.24\tabularnewline
WHO & Recreational & 2.0 & 113.14\tabularnewline
\bottomrule
\end{longtable}
